# Description
```{r, echo=FALSE, results="hide"}
file_names <-  list.files(path = "summariesdata", pattern = "savedRun", full.names = TRUE)
```

For the gompertz parameters, we have the following indexes:
```
gompertz[sex,race,par]

sex
    1:female
    2:male
race
    1:black
    2:white
    3:other
par
    1:B
    2:theta
```

Make three hazard rate functions that are easy to use.
If we want to add a linear predictor to these, the only thing we need to change is the `B`.
```{r}
h_t <- function(t, B, theta){
 return(B*exp(theta * t))
}

H_t <- function(t, B, theta){
 return( (B/theta) * (exp(theta*t) - 1) )
}

H_inv_t <- function(t, B, theta){
  return( log(t*(theta/B) + 1)/theta )
}
```




```{r}

# Use frspop dias, FRS1998, event_all  for now

a_file <- file_names[1]
a_file
```


```{r}
load("summariesdata/r_and_lambda_list.Rdata")
rlambdas <- reshape2::melt(r_and_lambdas_list)
normgammapars <- rlambdas[rlambdas$L1=="nhanes_pop_with_FRS_dias",]
normgammapars <- as.list(reshape2::dcast(normgammapars, .~L2 ))


alpha_1  <- normgammapars[["alpha_CLINIC"]]
alpha_2  <- normgammapars[["alpha_HOME"]]   
beta_1   <- normgammapars[["beta_CLINIC"]]  
beta_2   <- normgammapars[["beta_HOME"]]    
lambda_1 <- normgammapars[["lambda_CLINIC"]]
lambda_2 <- normgammapars[["lambda_HOME"]]  
m_1      <- normgammapars[["m_CLINIC"]]     
m_2      <- normgammapars[["m_HOME"]]       

```

Let's look at some people from the nhanesFRS 
```{r}

load("../Data_cleaned/nhanes_cleaned_lists.RData")
cleaned_nhanes_list <- list_nhanesFRS
sys_or_dias <- "dias"

x <- cleaned_nhanes_list[[sys_or_dias]]
xhome <- cleaned_nhanes_list[[paste0(sys_or_dias,"_home")]]

T <- cleaned_nhanes_list[["T"]]
nj <- cleaned_nhanes_list[["nj"]]
age <- cleaned_nhanes_list[["age"]]

normgammaparams     <- cleaned_nhanes_list[[paste0(sys_or_dias,"_params")]]$estimates
normgammaparams_home<- cleaned_nhanes_list[[paste0(sys_or_dias,"_home_params")]]$estimates

black <- cleaned_nhanes_list[["black"]]
white <- cleaned_nhanes_list[["white"]]
other <- cleaned_nhanes_list[["other"]]

racematrix <- t(t(data.frame(black, white, other) ) * c(1L,2L,3L))
# 1-black
# 2-white
# 3-other
race <- apply(racematrix, 1, max)

# 1-female
# 2-male
male <- cleaned_nhanes_list[["male"]]
sex <- as.integer(male) + 1L


# Make an inits function for stan
# Use the max rowsd for the tauis starting point
sds1 <- max(apply(x, 1, sd))
sds2 <- max(apply(xhome, 1, sd))
tauis <- cbind(rep(1/sds1^2, nrow(x)) , rep(1/sds2^2, nrow(x)))

M_i <- rowMeans( (as.matrix(x)+as.matrix(xhome))/2 )
Delta_i <- rowMeans( (as.matrix(x)-as.matrix(xhome))/2 ) 
mean_abs_delta <- mean( Delta_i )

FRS <- cleaned_nhanes_list[["FRS.1998"]]
```

Now to make the hazard rate:

```{r}
load(a_file) #loads the variable `resultslist` from one of the Rdata files
item <- resultslist
sumtable <- item$summary

estimates <- sumtable[,"mean"]

beta     <- estimates[1:4]
gompertz <- aperm(array(estimates[-c(1:4)], c(2,3,2)))

dimnames(gompertz) <- list( c("female", "male"),
                            c("black", "white", "other"),
                            c("B", "theta"))
                            
gompertz
```

```{r}

i <- 2 # number of the person in our dataset

df <- data.frame()

# For the first 1000 people
for(i in 1:1000 ){
  
  B     <- gompertz[sex[i],race[i],'B']
  theta <- gompertz[sex[i],race[i],'theta']
  
  Delta <- Delta_i[i]
  M <- M_i[i]
  tau_C <- tauis[i,1]
  tau_H <- tauis[i,2]
  
  FRS_i <- FRS[i]
  
  
  B_star <- B * 
    exp(
    beta[1] * ( abs(Delta)- mean_abs_delta)+
    beta[2] * (FRS_i)+
    beta[3] * (tau_H - (alpha_2/beta_2))+
    beta[4] * (tau_C - (alpha_1/beta_1))
    )
  
  
  #Ages:
  t <- 18:100
  model_h_t <- h_t(t, B=B_star, theta=theta)
  model_H_t <- H_t(t, B=B_star, theta=theta)

  
df <- rbind(df, data.frame(t= t, 
                           i=i,
                           h_t= model_h_t ,
                           H_t= model_H_t ,
                           sex=sex[i],
                           race=race[i],
                           Delta=Delta,
                           M=M,
                           tau_C=tau_C,
                           tau_H=tau_H,
                           FRS=FRS_i
                           )
                          )
                           
  
}

head(df)

```

```{r}
library(ggplot2)

cat("
sex
    1:female
    2:male
race
    1:black
    2:white
    3:other
")


p <- ggplot(data=df)
p <- p + aes(x=t, y=h_t, group=i, colour=as.factor(race))
p <- p + geom_line()
p <- p + facet_grid(sex ~ race)
p
```

# Conclusion
Finally, here is the Nelson-Aalen estimate of the hazard rate, overlaid on the $H(t)$ implied by our for the parameter values of the of the first 1000 people in the nhanes FRS population.
```{r}

library(survival)
event <- cleaned_nhanes_list$eventall
HT_nelaalen <- basehaz(coxph(Surv(age+T,event)~1 + strata(race, sex) ))


p <- ggplot(data=df)
p <- p + aes(x=t, y=H_t, group=i)
p <- p + geom_line()
p <- p + geom_point(aes(x=HT_nelaalen$time, y=HT_nelaalen$hazard, color=strata), data=HT_nelaalen)
p


```
